Sperm should evolve to make female meiosis fair

### 2014-05-20 writeup_and_notes/pnasSubmission/PNASSubmission.TEX
fixed minor tyos in significance statement and push PNAS folder

%%!TEX TS-program = latex
\documentclass[12pt,letterpaper]{article}
\usepackage{epsfig}                 
\usepackage[authoryear]{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{psfrag}
\usepackage{mathabx}
\renewcommand{\baselinestretch}{1.6}
\large
\pagenumbering{arabic}
\usepackage[usenames,dvipsnames]{color}
\usepackage{fullpage}
\bibliographystyle{genetics}
\usepackage{multirow}
\newcommand{\ye}{\hat{y}}
\newcommand{\xe}{\hat{x}}
\usepackage{color}
\usepackage[normalem]{ulem}  
	\newcommand{\gc}[1]{{ \color{red} #1}}
	\newcommand{\yb}[1]{{ \color{blue} #1}}

%Why do animals allow male control of female meiosis?
%Why does female meiosis allow the opportunity for exploitation by selfish sperm?
%sperm and egg genomes both favour the peaceful resolution of female meiotic drive
%strange bed fellows

%\title{Why does female meiosis allow the opportunity for exploitation by selfish sperm? BETTER TITLE?}
\title{
%Co-operation between the sexes trumps selfish control of female  meiotic drive. 
%Or 
Sperm should evolve to make female meiosis fair.
%Sperm do not evolve to collaborate in female meiotic drive. 
%Or Why does female meiosis allow the opportunity for exploitation by selfish sperm? 
}

\author{Yaniv Brandvain$^1$ \\ email: ybrandvain@gmail.com  \and Graham Coop$^2$ \\ email: gmcoop@ucdavis.edu }

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{}
\rhead{}
\renewcommand{\headrulewidth}{0.0pt}
\rfoot{}
\cfoot{\thepage}
\date{}
\bibliographystyle{plain}
\begin{document}
\maketitle
\begin{center}
$^1$Department of Plant Biology, \\ University of Minnesota --€“ Twin
Cities. St. Paul MN, 55108 \\
$^2$Center for Population Biology \& Department of Evolution and Ecology \\ University of California -- Davis. Davis, CA, 95616
\end{center}

%FIgures 
{\bf Abstract:}
Genomic conflicts arise when an allele gains an evolutionary advantage at a cost to organismal fitness. 
%Because only one of four products of female meiosis is transmitted to offspring, 
O\"{o}genesis is inherently susceptible to such conflicts
because alleles compete to be the  product of female meiosis transmitted to the egg. 
%(meiotic drive).  
Alleles that distort meiosis in their favor (i.e. meiotic drivers) often decrease organismal fitness,
	and therefore indirectly favor the evolution of mechanisms to suppress meiotic drive. % drive the rapid evolution of suppression mechanisms, and
In this light, many facets of  o\"{o}genesis and gametogenesis have been 
	interpreted as mechanisms of protection against genomic outlaws. 
Why then is female meiosis often left uncompleted until after
fertilization in many animals -- 
	potentially providing an opportunity for sperm alleles to
        meddle with its outcome and help like-alleles drive in heterozygous females? %and facilitate meiotic drive? 
The population genetic theory presented herein suggests that
	sperm nearly always evolve to increase the fairness of female
        meiosis in the face of genomic conflicts. %%felt this
                                %%might be an important addition as
                                %%they will spread if no cost to organism
These results are consistent with current knowledge of sperm-dependent meiotic drivers 
(loci whose distortion of female meiosis depends on sperm genotype),
	and suggest that the requirement of fertilization for the
        completion of female meiosis
        %, and variation in stage of meiosis that
        %fertilization occurs at 
        potentially represents a mechanism employed by females 
	 to ensure a fair meiosis. %prevent meiotic drive.
\newpage

\section*{Introduction}
Despite the apparent unity of the organism, `selfish' alleles can
gain an evolutionary advantage at a cost to  individual fitness
\citep{Burt2006}, often by exploiting meiosis and gametogenesis.
Because only one of the four products of female meiosis is transmitted to the egg, female meiosis is particularly vulnerable to such exploitation \citep{Sandler1957,Pardo-ManuelDeVillena2001a}. 
An allele that biases female meiosis in its favor (i.e. a meiotic driver), may increase in frequency even if it entails a pleiotropic fitness cost \citep{Prout1973}, generating a genetic conflict between the success of the driver and organismal fitness.
Meiotic drivers observed in both plants
\citep{Buckler1999,Fishman2005,Fishman2008}, and animals
\citep{Agulnik1990,Wu2005,Pardo-ManuelDeVillena2001c} highlight this
conflict -- the selfish benefits and the associated
pleiotropic fitness costs of drive sustain a balanced polymorphism
\citep{Prout1973}, 
and often generate ongoing evolutionary escalations of drive suppressors and enhancers \citep{Dawe1996,Fishman2008}. 
The threat of meiotic drive to organismal fitness is potentially so
severe that many basic properties of
meiosis and o\"{o}genesis, including the initial genome doubling in
meiosis I \citep{Haig1991}, arrested female meiosis
\citep{Mira1998}, the structure of centromere machinery \citep{Malik2002a,Malik2009}, and sex differences in the recombination rate \citep{Haig2010,Brandvain2012} 
have perhaps evolved to enforce fairness by disrupting meiotic drive \citep{Rice2013}.

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \citep{Masui_book}, 
	creating ample opportunity for interaction between the sperm and
	female meiotic machinery (note that the alternation of generations in
	likely precludes this interaction in plants). 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in heterozygous females %female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for exploitation by such selfish systems.  
Why then is the requirement of fertilization to complete female meiosis so ubiquitous?

It is certainly not the case that animals are mechanistically incapable of circumventing this requirement.
The stage at which female meiosis requires fertilization varies widely across taxa, and in a few animals  
	female meiosis is complete upon ovulation (that is, fertilization is not required for female meiosis \citep[e.g. Table 1 in 
][]{Masui_book}).

It is also not the case that sperm is mechanistically incapable of influencing female meiosis.
In \emph{C. elegans}, premature deployment of the aster, a vital component of mitotic
machinery provided by the sperm, disrupts MII meiotic segregation
in the egg, leading to a triploid zygote \citep{McNally2012}. 
Thus, there is direct mechanistic evidence that sperm can influence female meiosis.
 Additionally, genetic evidence from the \emph{In} and \emph{Om}  drive systems in mice suggests that the 
 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages
52-54 in][]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
% Both of these female drive systems operate by distorting the second meiotic division, 
% and in both systems the outcome of female meiosis depends the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
%While such genetic evidence is suggestive, we note the difficulty in
%ruling out the alternative hypothesis of early selection on zygotes \citep[pages
%52-54][]{Burt2006}.

We develop population genetic models exploring the evolution of sperm influence on female meiotic drive. %, focussing on models in which `self-promoting' alleles in sperm facilitate drive of like genotypes during gametogenesis in heterozygous females. 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
%In this article we explore through simple population genetic models the consequences of alleles that influence the outcome of female meiosis.
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such model universally reveal that sperm-acting drive modifiers are favored only if the suppress drive. 
These results highlight the fact that the interests of sperm and maternal  genomes' are often aligned, as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Therefore, there is little selective benefit to females in preventing sperm to influence female meioses.
Rather, features of o\"{o}genesis  may evolve to facilitate an interaction with sperm %with female meiosis
	as yet another mechanism to enforce an equitable transfer of genetic material.

\section*{Results}

\subsection*{ Invasion of the population by a driving allele that promotes itself.}
%%%%%%%%%%%%%%%%%%%%%%%%%
% PLEIOTROPY MODEL [self-promoting]
%%%%%%%%%%%%%%%%%%%%%%%%%

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype \cite[e.g. ][ and see Model 1 in the Appendix for more details]{Ubeda2004}. %The timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
%	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
%	so $d$ is bounded to be $<0.83$ \citep[see ][ for discussion]{Buckler1999}.
	%YB - MOVED THIS TO THE APPENDIX
          To depict a case
of a self-promoting meiotic driver,  we modify this standard model such 
	that the driver is only effective when fertilized by a sperm carrying that allele (see Figure \ref{Eggsperm}A and Model 2 in the Supplementary Worksheet). 
We then identify the conditions allowing for the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors. 
We conclude with an analysis of a related model (Model 3) - in which
drivers influence their transmission in females via paternal genotype,
rather than sperm haplotype.

For comparative purposes, we briefly present the standard drive model %as Model 1 in the Appendix, 
%equations \ref{eqn.A1}-\ref{eqn.A3}, 
	%and refer readers to the previous literature on this model 
	\citep[see e.g. ][ for additional results]{Prout1973,Ubeda2004}. 
Assuming the driving allele is deleterious in both sexes, but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$), a driver cannot fix, and
will be maintained as a protected polymorphism \citep{Prout1973}. 
The parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Eggsperm}B. %, and it can fix in the colored regions. 
When the allele is maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
a self-promoting driver specifically creates low fitness drive homozygotes 
	by uniting driving female gametes with sperm enabling that drive.
        % zygotes when it drives. in heterozygotes. 
%, and therefore there is a drive-associated fitness cost even when rare. 
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	%, the selection against the drive homozygote, 
	is less than $(2 d - 1)/(4 d)$. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Eggsperm}B).

When a self-promoting driver does spread, 
	it spends much time at low frequency, 
	because the paucity of complimentary sperm compromises its ability to drive. 
However, once relatively common, it  rapidly achieves fixation due to its
	positive frequency dependent behavior (Figure \ref{Eggsperm}B.1).  
%This lack of a balanced polymorphism %(Figure \ref{Invasion_space}) 
%	precludes the evolution of an
%	allele that suppresses this form of meiotic drive in females.
The positive frequency dependence of a self-promoting driver can also
	induce bistability in its spread -- some values of $s$ 
	allow the fixation of this driver when common, but 
	preclude its invasion when rare (Equation \ref{ineqA} and Figure \ref{Eggsperm}B). 
In this case, the driver will be fixed if it reaches a frequency greater than some threshold 
	(approximated in Equation \ref{bistab_homozy_approx} and  presented exactly in 
	Figure \ref{Bistab_homozyg_cost_fig}) and lost otherwise. 
For most parameters, this threshold is likely too high to be reached by drift, 
	and therefore  the fate of self-promoting drivers is determined
	by the more restrictive invasion criteria rather than the fixation criteria.

%I SUBSTANTIALLY REWROTE THIS
%The positive frequency dependence of a self-promoting driver can
%induce bistability in its spread.
%OLD GRAHAM WORDS
 %, i.e. there is a critical frequency
%above which it can spread quickly to fixation and below which it moves towards loss (an
%unstable equilibrium frequency). In 
%the appendix, we develop a simple
%approximation to this bistability. 
%When $s$ is too large to allow a rare driver to increase in frequency, but small enough to allow a common driver to fix (the inequality in Equation \ref{ineqA}),
%there is  an unstable equilibrium frequency, $f^*_{B\text{ recessive}}$  
%	above which it can spread quickly to fixation and below which it moves towards loss 
%	($f^*_{B\text{ recessive}}$ is approximated in Equation \ref{bistab_homozy_approx}.  
%	We compare this approximation to the exact result in Figure \ref{Bistab_homozyg_cost_fig}).	
%OLD GRAHAM WORDS
%This bistability occurs in the region of parameter space in Figure
%\ref{Invasion_space}. In Supplementary Figure
%\ref{Bistab_homozyg_cost_fig} we further explore this bistability and
%show that the unstable equilibrium frequency is high across much of
%this parameter space, and so in practice the self promoting allele
%only invade in the orange space. (In the appendix we develop a simple
%approximation to the unstable equilibrium frequency
%\eqref{bistab_homozy_approx} that is also shown in  Supplementary Figure
%\ref{Bistab_homozyg_cost_fig}).
%This unstable equilibrium frequency is high across much of
%	parameter space, likely too high to be reached by drift, 
%	and therefore  the fate of self-promoting drivers  will  likely be determined
%	by the more restrictive invasion criteria rather than the fixation criteria.

%OLD YANIV WORDS
% Compared to the invasion criteria, the parameter space
% allowing the fixation of the allele if it is above a cutoff frequency is much broader -- 
% 	a common, recessive, self-promoting driver fixes if
%         $s<(2d-1)/2$ approximately. 
% Therefore, when $(2 d - 1)/(4 d)<s<(2d-1)/(2)$, 
% 	a recessive, self-promoting driver will deterministically fix if drift, mutation, 
% 	or migration pressure bring its frequency above $(1-2d+4ds)/(2s(2d-1))$, 
% 	but it will be lost otherwise.
Inclusion of a heterozygous fitness cost (i.e. $w_{AB}=1-s_h$)
	further constrains the evolution of a self-promoting driver. 
In fact, with any heterozygous fitness cost, a rare self-promoting
	driver is always selected against. 
However, this case also displays bistability -- 
	when $s$ is sufficiently small (Equation \ref{sForHetFix}) this allele fixes deterministically if its 
	frequency exceeds some threshold  (Equation \ref{thresh2}, exact results in 
	Figure \ref{bistable_additive}).
%	$f_B^*$
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive})
%While a standard female driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
%	a rare self-promoter pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
%This frequency-independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Below a cutoff frequency the self-promoter driver deterministically decreases in frequency 
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive}). 
%Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
%	make drive effective enough to overcome the associated fitness cost.
This bistability prevents self-promoting drivers from invading 	
	reasonably sized populations, and assures that if they do invade, they will rapidly fix.
Our model therefore predicts that self-promoting drivers will not be
	observed as stable polymorphisms in natural populations. 
This lack of a balanced polymorphism %(Figure \ref{Invasion_space}) 
	precludes the evolution of an
	allele that suppresses this form of meiotic drive in females.

Although the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggests a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm), rather than sperm haplotype.
This paternal-genotype dependent model (Model 3) requires one additional parameter, as we exchange $d$ in the sperm dependent case for $d_{het}$ and $d_{hom}$ which describe the transmission of the drive allele in a heterozygous female mating with males heterozygous and homozygous for the self-promoting drive allele, respectively.  
Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and usually fixes when it invades.
	% fixes when $s$ is less than $d_{hom} - 1/2$.
However, when the distorting effect of genotype displays strong dominance in its
	effect on female meiosis ($d_{het}$ is close to $d_{hom}$), 
	there is a a narrow sliver of parameter space sustains a polymorphism
	when the cost of the drive is recessive
	(see Figure \ref{Effect_of_dominance}, and Equation \ref{polymale}).  
%Presumably this zone of polymorphism comes about because positive frequency 
%	dependence diminishes with the dominance of the male effect driver. 
%As an analytical example, consider the case of male-genotype dependent deleterious recessive female driver (assuming, as above, $w_{BB}=1-s$, and $w_{AA}=w_{AB}=1$). 
%Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and fixes when $s$ is less than $d_{hom} - 1/2$.
%Intermediate values of $s$, which represent a small proportion of parameter space can maintain a polymorphism. 
While mathematically interesting, it does not seem particularly realistic to think that the
	effect of the drive allele would be dominant in its action through the
	male genotype, while the cost would be recessive. 
Therefore, although Model 3 can sustain a polymorphism, 
	the lack of biological reality underlying the small portion of parameter values 
	required for this polymorphism make us doubt it general applicability. 
%Thus we regard this case as interesting but not very general.

@@

\begin{figure}
\includegraphics[width = 0.8 \textwidth]{Figures/selfprom.eps}
\caption{{\bf{A.}} A visual depiction of our model of 'self-promoting' driver. 
	Transmission probabilities for alleles through female
 	 meiosis depend on sperm genotype. 
	 The non-driving $A$-allele, and self-promoting $B$-allele
	 are represented by unfilled and filled circles, respectively. 
	 {\bf{B.}} Evolution of a self-promoter and standard driver. 
	 	Assuming that the fitnesses of drive homozygotes and 
		heterozygotes are $1-s$ and $1$, respectively. 
	 	\emph{Main figure:} Boundary conditions for the invasion 
		and fixation of self-promoting and standard meiotic drivers, 
		with drive coefficient, $d$. Colored regions depict exact results, 
		while lines  represent analytical approximations. 
	 \emph{B1:}  Trajectories of sperm-dependent female drive
		 %alleles that can invade the population, 
		 each allele has $s =0.1$ against the
		 homozygotes.  The drive coefficient is denoted by color. 
 }   
\label{Eggsperm}
\end{figure}

%\begin{figure}
%	\rotatebox{90}{
%\includegraphics[width = 0.5
   %       \textwidth]{Figures/Sperm_egg_cartoon_paper_Fig1.eps}}
%\caption{transmission probabilities for alleles through female
%  meiosis depend on sperm genotype. 2 allele model. \gc{merge with 2,
%    rewrite both captions}}  
%\label{Eggsperm_2_allele_cartoon}
%\end{figure}

% (we note that the timing of fertilization relative to female meiosis places another constraint on $d$, for example, when drive takes place druing the second meitoic division
%	it requires an uneven number of crossovers between the centromere and the drive locus \citep[see ][ for discussion]{Buckler1999}).

>>>>

%% PNAStwoS.tex
%% Sample file to use for PNAS articles prepared in LaTeX
%% For two column PNAS articles
%% Version1: Apr 15, 2008
%% Version2: Oct 04, 2013
\bibliographystyle{pnas}
%% BASIC CLASS FILE
\documentclass{pnastwo}

%% ADDITIONAL OPTIONAL STYLE FILES Font specification

%\usepackage{pnastwoF}

\usepackage{amsmath}
%% OPTIONAL MACRO DEFINITIONS
\def\s{\sigma}
%%%%%%%%%%%%
%% For PNAS Only:
\url{www.pnas.org/...}
\copyrightyear{20XX}
\issuedate{Issue Date}
\volume{Volume}
\issuenumber{Issue Number}
%\setcounter{page}{2687} %Set page number here if desired
%%%%%%%%%%%%

\begin{document}

\title{Sperm should evolve to make female meiosis fair}

\author{Yaniv Brandvain\affil{1}{Department of Plant Biology. University of Minnesota -- Twin Cities, St. Paul, MN, USA},
\and Graham Cooop\affil{2}{Department of Ecology and Evolution. University of California -- Davis. Davis, CA, USA}
}

\contributor{Submitted to Proceedings of the National Academy of Sciences
of the United States of America}

%%%Newly updated.
%%% If significance statement need, then can use the below command otherwise just delete it.
\significancetext{Female meiosis is the time during development where an allele can act to directly modify its representation in the next generation. 
The potential benefits to an allele that succeeds in this competition and collateral damage to organismal fitness are thought to have shaped basic features of female reproduction. 
However, little attention has been directed at the odd fact that fertilization by sperm is required to complete female meiosis in many species of animal. 
At first sight this seems strange as it seems to presents a dangerous opportunity for selfish alleles in sperm to distort the outcome of female meiosis in their own favor creating the opportunity for conflicts between male and female genomes. 
In this paper, through mathematical modeling, we show that in fact quite the opposite is true, sperm are usually selected to ensure the fairness of meiosis. 
Intuitively this comes about because selfish sperm have to immediately suffer any ill consequences of their actions in the resulting zygote, and so it is strongly in sperm's interests to keep genomes from gaming female meiosis. 
We hypothesize that the wide-spread requirement of fertilization to complete female meiosis may represent a long-term collaboration between male and female gametes to ensure fairness and the production of high quality gametes.}

\maketitle

\begin{article}
\begin{abstract}
{Genomic conflicts arise when an allele gains an evolutionary advantage at a cost to organismal fitness. 
%Because only one of four products of female meiosis is transmitted to offspring, 
O\"{o}genesis is inherently susceptible to such conflicts
because alleles compete to be the  product of female meiosis transmitted to the egg. 
Alleles that distort meiosis in their favor (i.e. meiotic drivers) often decrease organismal fitness,
	and therefore indirectly favor the evolution of mechanisms to suppress meiotic drive. In this light, many facets of  o\"{o}genesis and gametogenesis have been 
	interpreted as mechanisms of protection against genomic outlaws. 
Why then is female meiosis often left uncompleted until after
fertilization in many animals -- 
	potentially providing an opportunity for sperm alleles to
        meddle with its outcome and help like-alleles drive in heterozygous females? 
The population genetic theory presented herein suggests that
	sperm nearly always evolve to increase the fairness of female
        meiosis in the face of genomic conflicts.
These results are consistent with current knowledge of sperm-dependent meiotic drivers 
(loci whose distortion of female meiosis depends on sperm genotype),
	and suggest that the requirement of fertilization for the
        completion of female meiosis
        potentially represents a mechanism employed by females 
	 to ensure a fair meiosis.}
\end{abstract}

\keywords{meiosis | genomic conflict | meiotic drive | fertilization}

%\abbreviations{SAM, self-assembled monolayer; OTS, octadecyltrichlorosilane}

\dropcap{D}espite the apparent unity of the organism, `selfish' alleles can
gain an evolutionary advantage at a cost to  individual fitness
\cite{Burt2006}, often by exploiting meiosis and gametogenesis.
Because only one of the four products of female meiosis is transmitted to the egg, female meiosis is particularly vulnerable to such exploitation \cite{Sandler1957,Pardo-ManuelDeVillena2001a}. 
An allele that biases female meiosis in its favor (i.e. a meiotic driver), may increase in frequency even if it entails a pleiotropic fitness cost \cite{Prout1973}, generating a genetic conflict between the success of the driver and organismal fitness.
Meiotic drivers observed in both plants
\cite{Buckler1999,Fishman2005,Fishman2008}, and animals
\cite{Agulnik1990,Wu2005,Pardo-ManuelDeVillena2001c} highlight this
conflict -- the selfish benefits and the associated
pleiotropic fitness costs of drive sustain a balanced polymorphism
\cite{Prout1973}, 
and often generate ongoing evolutionary escalations of drive suppressors and enhancers \cite{Dawe1996,Fishman2008}. 
The threat of meiotic drive to organismal fitness is potentially so
severe that many basic properties of
meiosis and o\"{o}genesis, including the initial genome doubling in
meiosis I \cite{Haig1991}, arrested female meiosis
\cite{Mira1998}, the structure of centromere machinery \cite{Malik2002a,Malik2009}, and sex differences in the recombination rate \cite{Haig2010,Brandvain2012} 
have perhaps evolved to enforce fairness by disrupting meiotic drive \cite{Rice2013}.

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \cite{Masui_book}, 
	creating ample opportunity for interaction between the sperm and
	female meiotic machinery (note that the alternation of generations in
	likely precludes this interaction in plants). 
Therefore, a `green-bearded' \cite{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in heterozygous females %female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for exploitation by such selfish systems.  
Why then is the requirement of fertilization to complete female meiosis so ubiquitous?

It is certainly not the case that animals are mechanistically incapable of circumventing this requirement.
The stage at which female meiosis requires fertilization varies widely across taxa, and in a few animals  
	female meiosis is complete upon ovulation (that is, fertilization is not required for female meiosis (e.g. Table 1 in \cite {Masui_book}).

It is also not the case that sperm is mechanistically incapable of influencing female meiosis.
In \emph{C. elegans}, premature deployment of the aster, a vital component of mitotic
machinery provided by the sperm, disrupts MII meiotic segregation
in the egg, leading to a triploid zygote \cite{McNally2012}. 
Thus, there is direct mechanistic evidence that sperm can influence female meiosis.
 Additionally, genetic evidence from the \emph{In} and \emph{Om}  drive systems in mice suggests that the 
 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging (see pages 52-54 in \cite{Burt2006}), it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the genotype of the fertilizing sperm \cite{Agulnik1993,Wu2005}.

We develop population genetic models exploring the evolution of sperm influence on female meiotic drive. 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such model universally reveal that sperm-acting drive modifiers are favored only if the suppress drive. 
These results highlight the fact that the interests of sperm and maternal  genomes' are often aligned, as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \cite{Pomiankowski1993}).
Therefore, there is little selective benefit to females in preventing sperm to influence female meioses.
Rather, features of o\"{o}genesis  may evolve to facilitate an interaction with sperm
	as yet another mechanism to enforce an equitable transfer of genetic material.

\section{Results}
\subsection{Invasion of an allele that promotes its own drive}
In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype (e.g. \cite{Ubeda2004} and see Model 1 in the Appendix for more details). 
To depict a case of a self-promoting meiotic driver,  we modify this standard model such 
	that the driver is only effective when fertilized by a sperm carrying that allele (see Figure 1A and Model 2 	in the Supplementary Worksheet). 
We then identify the conditions allowing for the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors. 
We conclude with an analysis of a related model (Model 3) - in which
drivers influence their transmission in females via paternal genotype,
rather than sperm haplotype.

For comparative purposes, we briefly present the standard drive model 
	(see e.g. \cite{Prout1973,Ubeda2004} for additional results). 
Assuming the driving allele is deleterious in both sexes, but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because, when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$), a driver cannot fix, and
will be maintained as a protected polymorphism \cite{Prout1973}. 
The parameter space where the allele can invade but not fix is shown in white
        in Figure 1B. . 
When the allele is maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive (reviewed in \cite{Burt2006}).

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
a self-promoting driver specifically creates low fitness drive homozygotes 
	by uniting driving female gametes with sperm enabling that drive.
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	is less than $(2 d - 1)/(4 d)$. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure 1B).

When a self-promoting driver does spread, 
	it spends much time at low frequency, 
	because the paucity of complimentary sperm compromises its ability to drive. 
However, once relatively common, it  rapidly achieves fixation due to its
	positive frequency dependent behavior (Figure 1B.1).  
The positive frequency dependence of a self-promoting driver can also
	induce bistability in its spread -- some values of $s$ 
	allow the fixation of this driver when common, but 
	preclude its invasion when rare (Equation \ref{ineqA} and Figure 1B). 
In this case, the driver will be fixed if it reaches a frequency greater than some threshold 
	(approximated in Equation \ref{bistab_homozy_approx} and  presented exactly in 
	Figure S1) and lost otherwise. 
For most parameters, this threshold is likely too high to be reached by drift, 
	and therefore  the fate of self-promoting drivers is determined
	by the more restrictive invasion criteria rather than the fixation criteria.

Inclusion of a heterozygous fitness cost (i.e. $w_{AB}=1-s_h$)
	further constrains the evolution of a self-promoting driver. 
In fact, with any heterozygous fitness cost, a rare self-promoting
	driver is always selected against. 
However, this case also displays bistability -- 
	when $s$ is sufficiently small (Equation \ref{sForHetFix}) this allele fixes deterministically if its 
	frequency exceeds some threshold  (Equation \ref{thresh2}, exact results in 
	Figure S2).

This bistability prevents self-promoting drivers from invading 	
	reasonably sized populations, and assures that if they do invade, they will rapidly fix.
Our model therefore predicts that self-promoting drivers will not be
	observed as stable polymorphisms in natural populations. 
This lack of a balanced polymorphism
	precludes the evolution of an
	allele that suppresses this form of meiotic drive in females.

Although the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm (e.g. \cite{Joseph2004}) 
	suggests a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm), rather than sperm haplotype.
This paternal-genotype dependent model (Model 3) requires one additional parameter, as we exchange $d$ in the sperm dependent case for $d_{het}$ and $d_{hom}$ which describe the transmission of the drive allele in a heterozygous female mating with males heterozygous and homozygous for the self-promoting drive allele, respectively.  
Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and usually fixes when it invades.
However, when the distorting effect of genotype displays strong dominance in its
	effect on female meiosis ($d_{het}$ is close to $d_{hom}$), 
	there is a a narrow sliver of parameter space sustains a polymorphism
	when the cost of the drive is recessive
	(see Figure S3, and Equation \ref{polymale}).  
While mathematically interesting, it does not seem particularly realistic to think that the
	effect of the drive allele would be dominant in its action through the
	male genotype, while the cost would be recessive. 
Therefore, although Model 3 can sustain a polymorphism, 
	the lack of biological reality underlying the small portion of parameter values 
	required for this polymorphism make us doubt it general applicability.

@@

\subsection{Two locus models of sperm-dependent female drive}
Models 2 and 3, above, explored the dynamics of an allele that drove in females when signaled by a complementary signal in sperm.   
We complement this single-locus approach %of an allele with pleiotropic effects in both sexes, 
	with alternative models of two loci - one a female driver, 
	and the other, a sperm-acting allele which modifies the effect of drive upon fertilization. 
In this model, a female meiotic driver with no sperm-dependency 
	initially reaches drive-viability equilibrium (with two alleles
	$A$ and $B$ are the ancestral non-driver and driver alleles). 
Subsequently, a sperm-acting modifier of female meiotic drive arises at another locus.

We first assume that the modifier is tightly
        linked to the drive locus (effectively creating a third
        allele/haplotype at this locus) and arises on the drive-background. 
Tight linkage offers the best
        chance for a collaboration to evolve between a driver and
       a sperm-acting drive  enhancer, as recombination breaks up drive haplotypes \cite{Thomson1974,Charlesworth1978,Haig1991}. 
Additionally, tight linkage between female driver and sperm modifier is consistent with the nature of well characterized drive systems which are often maintained as polymorphic inversions with numerous linked modifiers \cite{Burt2006}. 
We conclude by analyzing models with alternative linkage relationship between driver and drive modifier - 
	in Model 5 the modifier arises tightly linked to the non-driving allele, 
	and in Model 6 it is unlinked to the driver.
